---
title: "Create a fish feeding EMODnet-Biology data product as NetCDF"
output: 
  html_document:
    code_folding: "show"
    df_print: 'tibble'
    number_section: yes
    code_download: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  collapse = TRUE,
  comment = "#>")
```

# Settings

There are four dependencies. The most important is the [RNetCDF](https://cran.r-project.org/web/packages/RNetCDF/index.html) package. This is a low-level R interface to the file format NetCDF designed by Unidata. Other packages to interact with NetCDF in R are [ncdf4](https://cran.r-project.org/web/packages/ncdf4/) and [tidync](https://cran.r-project.org/web/packages/tidync/).

The rest of libraries are used as helpers.

```{r libraries, warning=FALSE, message=FALSE}
library(RNetCDF)
library(readr)
library(dplyr)
library(glue)
```

# An example data set

This is **a fragment of a real data set**, adjusted to serve as a small example. The names of the columns are [Darwin Core terms](https://dwc.tdwg.org/terms)

```{r read}
load("../data/fish/dat_for_netcdf.Rdata")

dataset <- dat_for_netcdf
str(dataset)
names(dataset)[3] <- "axis"
```

# Arrange the data set

There is one variables of `av_bw_axis_scores`. This is numeric
I think it contains average  biomass in kg per sqkm?
It can be considered that this variable is defined by **four dimensions**: **Spatially**, these are `lon` and `lat`. **Timewise** is the column `year`.

The fourth dimension that defines the variable of interest here is the **axis**, which is shown in the column `axis`. There might be other info about these axes

## Technical constrain: Edit `year`

The `eventDate` **must be transformed to temporal amounts**. These are seconds, days, years etc since a certain date. There is a **helper** in the RNetCDF package: `uitinvcal.nc()`

In this demo, the units will be `days since 1970-01-01 00:00:00`.
Since the time is only provided as a year, we assume the first of january.

```{r date}
dataset$time <- utinvcal.nc(
  unitstring = "days since 1970-01-01 00:00:00" , 
  value = as.POSIXct(paste(dataset$year, "01", "01", sep = "-"), format = "%Y-%m-%d", tz = "UTC")
)

unique(dataset$eventDate)
unique(dataset$time)
```

## Get all possible combination of the dimensions

The key to transform a data frame to netcdf, is that **the data must be transformed into a 4D array**. To do that, **all the possible combinations of the dimensions must be attached**. This will coerce `NA` or empty values in the variable of interest as there will be no value for these combinations.

The base function `expand.grid()` allows to pass vectors and create a data frame with all the possible combinations of those vectors.

```{r coerce_into_array}
# First add an unique identifier by the combination of: 
# decimaLongitude, decimalLatitude, eventDate and AphiaID
dataset <- dataset %>%
  mutate(
    id = glue("{lon}-{lat}-{time}-{axis}")
  )

# Extract the unique and sorted values of the 4 dimensions
lon = sort(unique(dataset$lon))
lat = sort(unique(dataset$lat))
time = sort(unique(dataset$time))
axis = sort(unique(dataset$axis))


# Use expand.grid() to create a data frame with all the possible 
# combinations of the 4 dimensions
longer <- expand.grid(lon = lon, lat = lat, time = time, 
                     axis = axis, 
                     stringsAsFactors = FALSE)


# Define unique identifier again and merge the variable occurrenceStatus 
# with presences and absences

dataset <- dataset %>%
  select(id, av_bw_axis_scores)

longer <- longer %>% 
  mutate(
    id = glue("{lon}-{lat}-{time}-{axis}")
  ) %>%
  left_join(dataset) %>%
  select(-id)

# longer$date <- formatted_time <- as.POSIXct(longer$time *86400, origin = "1970-01-01", tz = "UTC")


# Save for later
write_csv(longer, "../data/derived/longer_fish.csv")

# Inspect
longer
```

## Turn data frame into a 4D array

**NetCDF is designed to host array-oriented scientific data** ([ref](https://docs.unidata.ucar.edu/netcdf-c/current/)). Transforming into an object of type `array` in R will allow the easiest way to add data into a NetCDF file.

In the demo dataset, we know that the **total length of each dimension** are:

-   lon: 29
-   lat: 51
-   time: 24
-   axis: 2

These lengths will be passed to the R function `array()` as a vector: `c(29, 51, 24, 2)`. We will also pass the variable of interest `longer$av_bw_axis_scores`. However, this variable must have as length the product of all lengths of all dimensions: **This is why we got all the combinations of the dimensions** via `expand.grid()`.

```{r create_array}
# The product of the lengths of all dimensions
29 * 51 * 24 * 2

# Is the same as the length of the variable of interest, including all 
# possible combinations of the dimensions even if this coerce NA's
length(longer$av_bw_axis_scores)

# Create array
array <- array(
  data = longer$av_bw_axis_scores,
  dim = c(length(lon), length(lat), length(time), length(axis))
)
```


```{r coerce_ices_into_array}
# First add an unique identifier by the combination of: 
# decimaLongitude, decimalLatitude, eventDate and AphiaID
dataset <- unique(dat_for_netcdf[,c("ices", "lon", "lat")]) %>%
  mutate(
    id = glue("{lon}-{lat}")
  )

# dataset <- dataset %>%
#   mutate(ices = substr(ices, 1, 2))

# Extract the unique and sorted values of the 4 dimensions
lon = sort(unique(dataset$lon))
lat = sort(unique(dataset$lat))



# Use expand.grid() to create a data frame with all the possible 
# combinations of the 4 dimensions
longer <- expand.grid(lon = lon, lat = lat, 
                     stringsAsFactors = FALSE)


# Define unique identifier again and merge the variable occurrenceStatus 
# with presences and absences

dataset <- dataset %>%
  select(id, ices)

longer <- longer %>% 
  mutate(
    id = glue("{lon}-{lat}")
  ) %>%
  left_join(dataset) %>%
  select(-id)



# Inspect
longer
```

## Turn data frame into a 4D array

**NetCDF is designed to host array-oriented scientific data** ([ref](https://docs.unidata.ucar.edu/netcdf-c/current/)). Transforming into an object of type `array` in R will allow the easiest way to add data into a NetCDF file.

In the demo dataset, we know that the **total length of each dimension** are:

-   lon: 29
-   lat: 51


These lengths will be passed to the R function `array()` as a vector: `c(29, 51)`. We will also pass the variable of interest `longer$ices`. However, this variable must have as length the product of all lengths of all dimensions: **This is why we got all the combinations of the dimensions** via `expand.grid()`.

```{r create_array_ices}
# The product of the lengths of all dimensions
29 * 51

# Is the same as the length of the variable of interest, including all 
# possible combinations of the dimensions even if this coerce NA's
length(longer$ices)

# Create array
array_ices <- array(
  data = longer$ices,
  dim = c(length(lon), length(lat))
)
```


# Transform into NetCDF

## Create a placeholder

The first step is to **create a netcdf file** that will work as a placeholder to add the data. We use the R package `RNetCDF`

```{r create_nc}
# Create nc file
nc <- create.nc("../data/derived/fish.nc") 
```

## Define dimensions

**Each dimensions has an homonymous variable assigned** along with number of attributes. This is a technical requirement. The dimensions must have the length of the unique values of each dimension.

Some extra attributes must be added to meet the [CF-Convention](http://cfconventions.org/).

In addition, the homonymous variable defining the dimensions **must have their own data**. The data are passed as a 1D vector to `var.put.nc()`, specifying in which variable has to be written.

### Longitude

```{r lon}
# Define lon dimension
dim.def.nc(nc, dimname = "lon", dimlength = length(lon)) 

# Define lon variable
var.def.nc(nc, varname = "lon", vartype = "NC_DOUBLE", dimensions = "lon")

# Add attributes
att.put.nc(nc, variable = "lon", name = "units", type = "NC_CHAR", value = "degrees_east")
att.put.nc(nc, variable = "lon", name = "standard_name", type = "NC_CHAR", value = "longitude")
att.put.nc(nc, variable = "lon", name = "long_name", type = "NC_CHAR", value = "Longitude")

# Put data
var.put.nc(nc, variable = "lon", data = lon) 

# Check
var.get.nc(nc, variable = "lon")
```

### Latitude

```{r lat}
# Define lat dimension
dim.def.nc(nc, dimname = "lat", dimlength = length(lat)) 

# Define lat variable
var.def.nc(nc, varname = "lat", vartype = "NC_DOUBLE", dimensions = "lat")

# Add attributes
att.put.nc(nc, variable = "lat", name = "units", type = "NC_CHAR", value = "degrees_north")
att.put.nc(nc, variable = "lat", name = "standard_name", type = "NC_CHAR", value = "latitude")
att.put.nc(nc, variable = "lat", name = "long_name", type = "NC_CHAR", value = "Latitude")

# Put data
var.put.nc(nc, variable = "lat", data = lat) 

# Check
var.get.nc(nc, variable = "lat")
```

### Time

```{r time}
# Define time dimension
dim.def.nc(nc, dimname = "time", dimlength = length(time)) 

# Define time variable
var.def.nc(nc, varname = "time", vartype = "NC_DOUBLE", dimensions = "time")

# Add attributes
att.put.nc(nc, variable = "time", name = "standard_name", type = "NC_CHAR", value = "time")
att.put.nc(nc, variable = "time", name = "long_name", type = "NC_CHAR", value = "Time")
att.put.nc(nc, variable = "time", name = "units", type = "NC_CHAR", value = "days since 1970-01-01 00:00:00")
att.put.nc(nc, variable = "time", name = "calendar", type = "NC_CHAR", value = "gregorian")

# Put data
var.put.nc(nc, variable = "time", data = time)

# Check
var.get.nc(nc, variable = "time")
```

### Axis

To define the axis dimension, there will be one dimension called `axis`, **with three variables assigned:**

-   `axis`: the actual axis name as a character
-   `taxon_name`: the scientific name of the taxa
-   `taxon_lsid`: the LSID (unique identifier) of the taxa

**These are technical requirements** both for ERDDAP and to meet the [CF-Convention](http://cfconventions.org/).

The dimension `aphiaid` will be used later to define the variable of interest. It is a better practice to use the AphiaID instead of, for instance, the scientific name in `taxon_name`, because **NetCDF works better with numeric data types than with characters.**

Note that **character variables in NetCDF require to be defined also by a dimension typically called "string"**. Its length is the **total number of characters** that can be hosted in the variable of type character. This is a requirement of NetCDF4.

#### axis

I had to make axis numeric so instead of MDS1 and MDS2 it is just 1 and 2

```{r axis}
# Define the axis and string80 dimensions
dim.def.nc(nc, dimname = "axis", dimlength = length(axis))
dim.def.nc(nc, dimname = "string80", dimlength = 80)

# Add axis variable and attributes 
var.def.nc(nc, varname = "axis", vartype = "NC_INT", dimensions = "axis")
axis_long_name = "feeding guild axis" 
att.put.nc(nc, variable = "axis", name = "long_name", type = "NC_CHAR", value = axis_long_name)
att.put.nc(nc, variable = "axis", name = "units", type = "NC_CHAR", value = "level")
# Put axis data
var.put.nc(nc, variable = "axis", data = as.numeric(substring(axis, 4,4)))

# Check
var.get.nc(nc, variable = "axis")
```

#### axis_name

```{r axis_name}
# Add axis name variable and attributes
var.def.nc(nc, varname = "axis_name", vartype = "NC_CHAR", dimension = c("string80", "axis"))
att.put.nc(nc, variable = "axis_name", name = "axis_name", type = "NC_CHAR", value = "feeding guild axis name")

# Put axis_name data
var.put.nc(nc, variable = "axis_name", data = c("scales between benthivores and planktivores", "scales between piscivores and planktivores/benthivores"))

# Check
var.get.nc(nc, variable = "axis_name")


```



### Coordinate Reference System

A non-dimensional variable will be defined to host all the info about the **Coordinate Reference System (CRS)**. This is useful for some GIS software.

**It is assumed that the CRS is WGS84**. If it was different: transform `decimalLatitude` and `decimalLongitude` to WGS84 before hand. See R package `sf`.

```{r crs}
# Define non-dimensional crs variable 
var.def.nc(nc, varname = "crs", vartype = "NC_CHAR", dimensions = NA)

# Add attributes
att.put.nc(nc, variable = "crs", name = "long_name", type = "NC_CHAR", value = "Coordinate Reference System")
att.put.nc(nc, variable = "crs", name = "geographic_crs_name", type = "NC_CHAR", value = "WGS 84")
att.put.nc(nc, variable = "crs", name = "grid_mapping_name", type = "NC_CHAR", value = "latitude_longitude")
att.put.nc(nc, variable = "crs", name = "reference_ellipsoid_name", type = "NC_CHAR", value = "WGS 84")
att.put.nc(nc, variable = "crs", name = "horizontal_datum_name", type = "NC_CHAR", value = "WGS 84")
att.put.nc(nc, variable = "crs", name = "prime_meridian_name", type = "NC_CHAR", value = "Greenwich")
att.put.nc(nc, variable = "crs", name = "longitude_of_prime_meridian", type = "NC_DOUBLE", value = 0.)
att.put.nc(nc, variable = "crs", name = "semi_major_axis", type = "NC_DOUBLE", value = 6378137.)
att.put.nc(nc, variable = "crs", name = "semi_minor_axis", type = "NC_DOUBLE", value = 6356752.314245179)
att.put.nc(nc, variable = "crs", name = "inverse_flattening", type = "NC_DOUBLE", value = 298.257223563)
att.put.nc(nc, variable = "crs", name = "spatial_ref", type = "NC_CHAR", value = 'GEOGCS[\"WGS 84\",DATUM[\"WGS_1984\",SPHEROID[\"WGS 84\",6378137,298.257223563]],PRIMEM[\"Greenwich\",0],UNIT[\"degree\",0.0174532925199433,AUTHORITY[\"EPSG\",\"9122\"]],AXIS[\"Latitude\",NORTH],AXIS[\"Longitude\",EAST],AUTHORITY[\"EPSG\",\"4326\"]]')
att.put.nc(nc, variable = "crs", name = "GeoTransform", type = "NC_CHAR", value = '-180 0.08333333333333333 0 90 0 -0.08333333333333333 ')
```

## Define the variable of interest

The variable to add the presence/absence data **must be defined by the four dimensions** considered in this demo. This is passed to `var.def.nc()` in the argument `dimensions` as **a vector containing the names of the dimensions.**

The values stating presences and absences are 1 and 0, hence the variable must be of type integer.

Some other attributes are added. E.g. `_FillValue` is a requirement for the CF-Convention stating what value will be used in case of NULL or NA. This is typically `-99999`. The attribute `long_name` is free text and it describes the variable.

```{r def_biomass}
# Create the biomass variable defined by the four dimensions
var.def.nc(nc, varname = "axis_score", vartype = "NC_DOUBLE", dimensions = c("lon", "lat", "time", "axis"))
av_bw_axis_scores_short_name = "non-metric multidimensional axis scores"
av_bw_axis_scores_long_name = "biomass-weighted mean axis scores from non-metric multidimensional analysis (NMDS) of fish stomach contents data" # explain what this is
# Add attributes
att.put.nc(nc, variable = "axis_score", name = "_FillValue", type = "NC_DOUBLE", value = -99999)
att.put.nc(nc, variable = "axis_score", name = "short_name", type = "NC_CHAR", value = av_bw_axis_scores_short_name)
att.put.nc(nc, variable = "axis_score", name = "long_name", type = "NC_CHAR", value = av_bw_axis_scores_long_name)
#explain what the score means
score_comment = "Axis 1 shows the difference between bethivores and planktivores. Axis 2 shows the difference between piscivores and planktivores/benthivores"
att.put.nc(nc, variable = "axis_score", name = "comment", type = "NC_CHAR", value = score_comment)


```

```{r def_ices, eval = F}
# Create the biomass variable defined by the four dimensions
var.def.nc(nc, varname = "ices_rect", vartype = "NC_CHAR", dimensions = c("lon", "lat"))
var.def.nc(nc, varname = "test", vartype = "NC_CHAR", dimensions = c("lon", "lat"))
# Add attributes

att.put.nc(nc, variable = "ices_rect", name = "long_name", type = "NC_CHAR", value = "ices_rectangle")


```

If there were a standard name for such a variable in the [CF-Convention](http://cfconventions.org/), this would be add to an attribute named `standard_name`.

### Add data from a 4D Array

As we previously created a 4D array, we can pass this directly to `var.put.nc()`

```{r put_array}
var.put.nc(nc, variable = "axis_score", data = array) 

# Check
var.get.nc(nc, variable = "axis_score")
```

```{r put_array_ices, eval = F}
var.put.nc(nc, variable = "ices_rect", data = array_ices) 

# Define the count vector to match the dimensions of the variable
count <- c(dim(array_ices))

# Write data to the NetCDF variable
var.put.nc(nc, variable = "test", data = t(array_ices))

var.put.nc(nc, variable = "test3", data = array_ices, count = dim(array_ices))
# Check
var.get.nc(nc, variable = "test")
```

## Global Attributes

NetCDF files can host **metadata in the form of global attributes**. These contain information such as the author, data of creation, affiliation, citation or license.

See below an example of how defining a number of global attributes and adding them programatically.

**Click on `code` to unfold**

```{r global_att, class.source = "fold-hide"}
attributes <- list(
  title = "Biomass-Weighted Fish Functional Feeding Traits for the Northeast Atlantic (1997-2020)",
  summary = "This product enables users to assess spatial and temporal change in fish functional feeding traits across the northeast Atlantic shelf seas. Values are based on non-metric multidimensional axis scores, weighted by biomass observed in otter trawl surveys",                       
  Conventions = "CF-1.8",
  # id = "",
  naming_authority = "emodnet-biology.eu",
  history = "", # github link here
  source = "", # github link here
  # processing_level = "",
  comment = "This product is the result of collating and analyzing fish stomach contents data from the North Atlantic and Arctic Oceans to classify fish into feeding guilds. The feeding guilds identified include planktivores, benthivores, bentho-piscivores, and piscivores. Non-metric multidimensional scaling (NMDS) analysis was used to provide a more nuanced understanding of different predator feeding traits in relation to others. The NMDS analysis presents axis scores that offer a bounded data perspective, as opposed to categorical, of different predator feeding traits. This analysis is part of the process to classify feeding guilds based on predator stomach contents data. See figure S4 in https://essd.copernicus.org/preprints/essd-2024-102/essd-2024-102-supplement.pdf for a pictorial depiction. Uses attributes recommended by http://cfconventions.org", #	Miscellaneous information about the data or methods used to produce it. I got this info from the essd paper (edited copilot summary of methods)
  # acknowledgment = "",
  license = "CC-BY",
  standard_name_vocabulary = "CF Standard Name Table v1.8",
  date_created = as.character(Sys.Date()),
  creator_name = "Murray Thompson",
  creator_email = "murray.thompson@cefas.gov.uk",
  creator_url = "www.cefas.co.uk",
  institution = "Centre for Environment, Fisheries and Aquaculture Science (Cefas)",
  project = "EMODnet-Biology",
  publisher_name = "EMODnet-Biology",                 
  publisher_email = "bio@emodnet.eu",                
  publisher_url = "www.emodnet-biology.eu",                  
  # geospatial_bounds = "",              
  # geospatial_bounds_crs = "",          
  geospatial_lat_min = min(lat),
  geospatial_lat_max = max(lat),
  geospatial_lon_min = min(lon),
  geospatial_lon_max = max(lon),
  time_coverage_start = "1997",            
  time_coverage_end = "2020",              
  # time_coverage_duration = "",         
  time_coverage_resolution = "years",       
  # uuid = "",                           
  sea_name = "North West Atlantic Shelf",                       
  # creator_type = "",                   
  creator_institution = "Centre for Environment, Fisheries and Aquaculture Science (Cefas )",            
  # publisher_type = "",                 
  publisher_institution = "Flanders Marine Institute (VLIZ)",        
  # program = "",                        
  contributor_name = "Oliver Williams, Kate Collingridge",               
  contributor_role  = "Project PI, dataset preparation",              
  geospatial_lat_units = "degrees_north",           
  geospatial_lon_units = "degrees_east",           
  # date_modified = "",               
  # date_issued = "",                    
  # date_metadata_modified   = "",       
  # product_version = "",            
  keywords_vocabulary = "http://vocab.nerc.ac.uk/collection/P21/current/MS11088/", 
  # platform  = "",              
  # platform_vocabulary = "",          
  instrument = "otter trawl",          
  instrument_vocabulary  = "https://vocab.nerc.ac.uk/collection/L22/current/NETT0183/",        
  # featureType = "Point",                  
  metadata_link = "https://data.cefas.co.uk/view/21771",                  
  references = "https://doi.org/10.14466/CefasDataHub.149 ; Thompson et al 2024 and https://doi.org/10.5194/essd-2024-102; Thompson et al in review",
  citation = "Murray et al (2025). Biomass-Weighted Fish Functional Feeding Traits for the Northeast Atlantic. (1997-2020).",
  acknowledgement = "European Marine Observation Data Network (EMODnet) Biology project (EMFF/2019/1.3.1.9/Lot 6/SI2.837974), funded by the European Union under Regulation (EU) No 508/2014 of the European Parliament and of the Council of 15 May 2014 on the European Maritime and Fisheries Fund"
)

# Define function that detects if the data type should be character of 
# integer and add to global attributes
add_global_attributes <- function(nc, attributes){
  
  stopifnot(is.list(attributes))
  
  for(i in 1:length(attributes)){
    if(is.character(attributes[[i]])){
      type <- "NC_CHAR"
    }else if(is.numeric(attributes[[i]])){
      type <- "NC_DOUBLE"
    }
    att.put.nc(nc, variable = "NC_GLOBAL", name = names(attributes[i]), type = type, value = attributes[[i]])
  }
  sync.nc(nc)
}

# Add attributes
add_global_attributes(nc, attributes)
```

## Wrapping up

Congratulations! The NetCDF file containing the product has been created. Now you can inspect the file.

```{r inspect}
sync.nc(nc)
print.nc(nc)
```

If everything is correct, close the file. Remember it was saved into `./data/derived/foo.nc`

```{r close}
close.nc(nc)
```


```{r netcdf4}
library(ncdf4)
nc2 <- nc_open("../data/derived/fish2.nc", write = TRUE)
names(nc2$var)

# Define the dimensions (assuming they already exist in the file)
lon_dim <- nc2$dim$lon
lat_dim <- nc2$dim$lat

# Define the new variable
new_var <- ncvar_def(name = "ices_rect", units = "units", dim = list(lon_dim, lat_dim),longname = "ices rectangle", prec = "char")


# Add the new variable to the NetCDF file
nc2 <- ncvar_add(nc2, new_var)
names(nc2$var)[4] <- "ices_rect"
ncvar_put(nc2, new_var, array_ices)

# Close the NetCDF file
nc_close(nc2)
```
